Emergence of the Southeast Asian islands as a driver for Neogene cooling

### 2020-04-08 Manuscript/Manuscript.tex
Updates from Overleaf

% other packages
\usepackage{xspace}

% self-defined commands
\newcommand{\degreesC}{\textdegree C\xspace}
\newcommand{\degrees}{\textdegree\xspace}
\newcommand{\dC}{$\delta^{13}$C\xspace}
\newcommand{\dO}{$\delta^{18}$O\xspace}
\newcommand{\SrSr}{$^{87}$Sr/$^{86}$Sr\xspace}
\newcommand{\OsOs}{$^{187}$Os/$^{188}$Os\xspace}
\newcommand{\permil}{\textperthousand\xspace}
\newcommand{\pCOtwo}{\textit{p}CO$_{2}$\xspace}
\newcommand{\COtwo}{CO$_{2}$\xspace}

@@

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{Please provide details of author contributions here.}
\authordeclaration{No competing interests to declare.}
\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: yuempark@berkeley.edu}

% At least three keywords are required at submission. Please provide three to five keywords, separated by the pipe symbol.
\keywords{Keyword 1 $|$ Keyword 2 $|$ Keyword 3 $|$ ...}

@@

Over geologic time-scales, \COtwo enters Earth's ocean--atmosphere system primarily via volcanism and metamorphic degassing, and leaves primarily through the chemical weathering of silicate rocks, with a smaller sink through organic carbon burial \cite{Kump1997a}. Chemical weathering delivers alkalinity and cations to the ocean which drives carbon sequestration through carbonate precipitation. Steady-state \pCOtwo is set at the \pCOtwo level at which \COtwo sinks are equal to sources. As \COtwo sinks are removed and \pCOtwo rises, temperature increases and the hydrological cycle is invigorated, causing other weathering sinks to increase until a new steady-state is achieved at higher \pCOtwo \cite{Kump1997a}. Conversely, as regions with high carbon sequestration potential emerge on Earth's surface, \pCOtwo decreases resulting in other weathering sinks to diminish leading to a new lower steady-state \pCOtwo concentration.

Topography, climate, and lithology all effect chemical weathering. High-relief regions lead to reaction-limited weathering regimes that are more sensitive to climate change \cite{Gabet2009a, West2012a}. In warm and wet regions, mineral dissolution kinetics are faster leading to enhanced chemical weathering \cite{West2012a}. Mafic rocks have higher Ca and Mg concentrations and dissolution rates than felsic rocks, and thus the potential to more efficiently sequester carbon through silicate weathering \cite{Dessert2003a}. These factors have lead to the proposal that arc-continent collisions within the tropical rain belt have been important in enhancing global weatherability, lowering atmospheric \pCOtwo, and initiating glacial climate over the past 520~m.y. \cite{Jagoutz2016a, Swanson-Hysell2017a, Macdonald2019a} and perhaps in the Neoproterozoic as well \cite{Park2019b}.

To estimate the decrease in steady-state \pCOtwo associated with the increase of subaerially-exposed land area in SEAI, we use the global spatially-resolved GEOCLIM model \cite{Godderis2017c}. GEOCLIM estimates changes in steady-state \pCOtwo associated with coupled changes in chemical weathering and climatology by linking a silicate weathering model to climate model runs at multiple \pCOtwo levels.

The silicate weathering component of GEOCLIM calculates \COtwo consumption resulting from silicate weathering for subaerially-exposed land. In previous versions of the model, silicate weathering was a function of temperature and runoff only, and all bedrock was assigned identical chemical compositions \cite{Godderis2017c}. More recent versions of GEOCLIM implement regolith development and soil shielding, which introduces a dependence on erosion (and therefore topographic slope) \cite{Maffre2018a}. While this introduction of regolith development into GEOCLIM is important for assessing the impact of tropical arc-continent collisions on \pCOtwo, the relatively high Ca+Mg concentration in arc rocks relative to other lithologies must also be considered. We therefore implement variable bedrock Ca+Mg concentration into GEOCLIM. The spatial distribution of lithologies is sourced from the Global Lithologic Map (GLiM) \cite{Hartmann2012a} and is represented by 6 categories: metamorphic, felsic, intermediate, mafic, carbonate, and siliciclastic sediment (see SI). Each land pixel is assigned these lithologic categories at a resolution of 0.1\degrees $\times$ 0.1\degrees. The Ca+Mg concentrations of felsic, intermediate, and mafic lithologies are assigned based on the mean of data compiled from EarthChem (\url{www.earthchem.org/portal}). Given that GLiM does not distinguish ultramafic lithologies, such rocks are grouped with mafic rocks. As a result, the Ca+Mg concentration is likely an underestimate in regions of obducted ophiolites, such that the estimated effect of these regions on changing steady-state \pCOtwo is conservative. The weathering of carbonate does not contribute to long-term \COtwo consumption and its Ca+Mg concentration is ignored. The Ca+Mg concentrations of metamorphic and siliciclastic sediment lithologies are more difficult to define, since their chemical composition is strongly dependent on protolith composition and, in the case of siliciclastic sediment, the degree of previous chemical depletion. We explore a range of feasible Ca+Mg concentrations for metamorphic rocks and siliciclastic sediment during calibration of the silicate weathering component.

The values of four parameters within the silicate weathering component that modify the dependence of silicate weathering on temperature, runoff, and regolith thickness are poorly constrained. Rather than prescribing values, we allow these parameters and the Ca+Mg concentration of metamorphic and siliciclastic pixels to vary within reasonable bounds (see SI). This results in 93,600 unique combinations of these six parameters. For each combination, we compute spatially-resolved long-term \COtwo consumption associated with Ca+Mg fluxes using present-day runoff, temperature, and slope. We sum computed \COtwo consumption over watersheds for which data-constrained estimates are available \cite{Gaillardet1999a, Moquet2018a}, then calculate the coefficient of determination ($r^{2}$) between computed and measured \COtwo consumption in each of these watersheds. After eliminating parameter combinations that result in low $r^{2}$, 5,381 parameter combinations remain. The resulting global \COtwo consumption of these model runs all overlap with independently-derived estimates of global \COtwo outgassing flux as they should with the long-term carbon cycle being at steady-state (see SI).

Having calibrated the silicate weathering component of GEOCLIM, we use it to estimate the decrease in steady-state \pCOtwo associated with SEAI emergence. For the climate model component, we use temperature and runoff from a subset of the GFDL CM2.0 experiments \cite{Delworth2006b} (see SI). The primary strength of these experiments for this analysis is that all non-\COtwo forcings are held constant at values representative of pre-industrial conditions, allowing the effect of changing \pCOtwo on climatology to be isolated.

To determine the position of SEAI paleoshorelines over the past 15~m.y., we use terrestrial and marine sedimentary deposits (Fig. \ref{fig:shoreline_growth}; see SI). The paleoshoreline data indicate that the Sunda-Banda Arc and New Guinea are primarily responsible for the increase in area since 15~Ma. Exhumation of the modern Sunda-Banda Arc is the result of ongoing arc-continent collision with the subducting Australian Plate \cite{Harris2006a}. Of the major islands in this arc, only Sumatra, Belitung, Bangka, Java, Bali, and Flores were subaerially exposed before 5~Ma. However, most of Sumatra and Java along with the non-volcanic islands of the Outer Banda Arc were elevated above sea level after 5~Ma \cite{Hall2013b}. In New Guinea, emergence in the Middle Miocene is associated with collision between the Melanesian Arc and Australia's distal margin \cite{Cloos2005a}, which drove exhumation of the Irian-Marum-April Ophiolite Belt. Exhumation accelerated over the past 4~m.y. in the New Guinea Central Range due to slab-breakoff and buoyant uplift, and in eastern New Guinea due to jamming of the north-dipping subduction zone \cite{Cloos2005a}. These tectonic drivers and others throughout the region lead to progressive emergence over the past 15~m.y. that accelerated following 5~Ma (Fig. \ref{fig:shoreline_growth}B). This trend mirrors broad cooling over the Neogene that resulted in the initiation of Northern Hemisphere ice sheets (Fig. \ref{fig:shoreline_growth}C). We also include changes in areas of presently-submerged continental shelves such as the Sunda Shelf that were previously exposed (see SI).

We use GEOCLIM to estimate \pCOtwo associated with the reconstructed subaerial extent of SEAI at ca. 15, 10, and 5~Ma (``paleo-SEAI'' scenarios; Fig. \ref{fig:scenario_pCO2}). Because we use a clmate model forced with modern geography, the position of the tectonic blocks remain fixed. Although there has been motion of these tectonic blocks since 15~Ma, they have remained within tropical latitudes such that this fixed scenario is a good approximation of the paleogeography (see SI). We also test an end-member scenario, in which all islands associated with arc-continent collision in the region are removed (``removed SEAI'' scenario).

Using the 5,381 unique parameter combinations, the ``paleo-SEAI'' scenarios resulted in 378--447~ppm \pCOtwo for 5~Ma, 435--532~ppm for 10~Ma, and 493--709~ppm for 15~Ma (Fig. \ref{fig:scenario_pCO2}). These results indicate a progressive decrease in \pCOtwo over the Neogene associated with the emergence of SEAI, and suggest that without this emergence, pre-industrial \pCOtwo would have been $\sim$493--709~ppm. Proxy-based estimates of the magnitude and trajectory of \pCOtwo change from the Miocene to the Pliocene are variable between techniques and associated assumptions underlying their interpretation \cite{Foster2017a}. The modeled \pCOtwo values for 15~Ma resemble the higher end of proxy-based \pCOtwo estimates for the early-mid-Miocene (see SI), indicating that the increase in subaerially-exposed land area and tectonic topography of SEAI is sufficient to explain long-term cooling of Earth's climate over the Neogene. The \pCOtwo threshold for Antarctic glaciation is estimated to be $\sim$750~ppm with that for Northern Hemisphere glaciation being significantly lower at $\sim$280~ppm \cite{DeConto2008a}. These modeled values of decreasing \pCOtwo are therefore consistent with the record of Neogene climate with Miocene ice sheets on Antarctica \cite{Sugden1995a} followed by Northern Hemisphere ice sheets developing in the Pliocene \cite{Haug2005a} as \pCOtwo subsequently decreased.

@@

Himalayan uplift \cite{Raymo1992a, Galy2007a} or a decrease in volcanic outgassing \cite{Berner1983a} have also been proposed as drivers for Neogene cooling. Rising marine \SrSr has been associated with increasing weathering of the high \SrSr Himalayan lithologies \cite{Raymo1992a}. This increase in \SrSr since ca. 35~Ma is followed by a decrease in slope ca. 15~Ma that has been attributed to exhumation of relatively low \SrSr and high \OsOs Lesser Himalayan lithologies \cite{Colleps2018a}, but could also be at least partially driven by the exhumation of mafic and organic-rich forearc sediments in SEAI during arc-continent collision. Enhanced organic matter burial in the Bay of Bengal due to high sedimentation rates could also have contributed to Neogene cooling \cite{Galy2007a}.

It has been argued that enhanced weathering in regions of active uplift predicts an increase in global alkalinity fluxes. However, global alkalinity delivery from silicate weathering needs to be nearly constant to keep the long-term carbon cycle in steady-state \cite{Kump1997a}. Enhanced weathering in a region such as SEAI is compensated by a decrease in weathering elsewhere. Global alkalinity delivery from silicate weathering does not change, but occurs more efficiently and thereby at lower \pCOtwo. Given that carbonate weathering is disconnected from the long-term carbon-cycle mass balance, changes in carbonate accumulation through time \cite{Si2019a} could be driven by changes in carbonate weathering.

Coupled geological constraints and modeling experiments demonstrate that SEAI has been a growing hot spot for carbon sequestration from the Miocene to present. Changes in volcanic outgassing and paleogeography elsewhere on Earth, particularly in the Himalaya and Central America, would have also effected geological carbon sources and sinks. Such changes are likely to have induced complex feedbacks on many Earth surface processes. Yet, not only does the history of SEAI emergence coincide with Neogene cooling and the onset of Northern Hemisphere glaciation, but our coupled weathering-climate model also indicates that the associated steady-state \pCOtwo change is sufficient to explain the magnitude of cooling. These results highlight that the Earth's climate state is particularly sensitive to changes in tropical paleogeography.

\section*{Guide to using this template on Overleaf}

\begin{figure}
    \centering
    \includegraphics[width=0.9\linewidth]{Figures/shoreline_growth.pdf}
    \caption{The emergence of the Southeast Asian islands (also referred to as the Maritime Continent in climate science literature) from the mid-Miocene to present. Past shorelines are shown in A with associated land area summarized in B. A significant increase in area over the past 5 million years is coincident with cooling and the onset of Northern Hemisphere glaciation as reflected in the benthic oxygen isotope record \cite{Zachos2008a} shown in C.}
    \label{fig:shoreline_growth}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=1\linewidth]{Figures/scenario_pCO2.pdf}
    \caption{Steady-state \pCOtwo estimates from GEOCLIM for the various scenarios discussed in the text. For each of the seven scenarios, each point represents an estimate from one of the 5,381 unique parameter combinations that resulted in reasonable total global \COtwo consumption and most closely matched estimates of present-day \COtwo consumption in 80 watersheds around the world (see SI). The box encloses the middle 50\% of the \pCOtwo estimates (i.e. the interquartile range), and the notch represents the median with its 95\% confidence interval. The whiskers extend to the 2.5 and 97.5 percentile values. Glaciation thresholds \cite{DeConto2008a} are shown on the x-axis.}
    \label{fig:scenario_pCO2}
\end{figure}

\subsection*{Author Affiliations}

x

\begin{figure}%[tbhp]
\centering
\includegraphics[width=.8\linewidth]{Figures/regolith_schematic.pdf}
\caption{Placeholder image of a frog with a long example legend to show justification setting.}
\label{fig:frog}
\end{figure}

\begin{SCfigure*}[\sidecaptionrelwidth][t]
\centering
\includegraphics[width=11.4cm,height=11.4cm]{Figures/regolith_schematic.pdf}
\caption{This legend would be placed at the side of the figure, rather than below it.}\label{fig:side}
\end{SCfigure*}

\begin{figure*}[bt!]
\begin{align*}
(x+y)^3&=(x+y)(x+y)^2\\
       &=(x+y)(x^2+2xy+y^2) \numberthis \label{eqn:example} \\
       &=x^3+3x^2y+3xy^3+x^3. 
\end{align*}
\end{figure*}

\begin{table}%[tbhp]
\centering
\caption{Comparison of the fitted potential energy surfaces and ab initio benchmark electronic energy calculations}
\begin{tabular}{lrrr}
Species & CBS & CV & G3 \\
\midrule
1. Acetaldehyde & 0.0 & 0.0 & 0.0 \\
2. Vinyl alcohol & 9.1 & 9.6 & 13.5 \\
3. Hydroxyethylidene & 50.8 & 51.2 & 54.0\\
\bottomrule
\end{tabular}

\addtabletext{nomenclature for the TSs refers to the numbered species in the table.}
\end{table}

\matmethods{Please describe your materials and methods here. This can be more than one paragraph, and may contain subsections and equations as required. Authors should include a statement in the methods section describing how readers will be able to access the data in the paper.

\subsection*{Subsection for Method}
Example text for subsection.
}

>>>>

% other packages
\usepackage{xspace}
\urlstyle{same}

% self-defined commands
\newcommand{\degreesC}{\textdegree C\xspace}
\newcommand{\degrees}{\textdegree\xspace}
\newcommand{\dC}{$\delta^{13}$C\xspace}
\newcommand{\dO}{$\delta^{18}$O\xspace}
\newcommand{\SrSr}{$^{87}$Sr/$^{86}$Sr\xspace}
\newcommand{\OsOs}{$^{187}$Os/$^{188}$Os\xspace}
\newcommand{\permil}{\textperthousand\xspace}
\newcommand{\pCOtwo}{\textit{p}CO$_{2}$\xspace}
\newcommand{\COtwo}{CO$_{2}$\xspace}
\newcommand{\SI}{\textit{SI Appendix}\xspace}
\newcommand{\MM}{\textit{Materials and Methods}\xspace}

@@

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{Please provide details of author contributions here.}
\authordeclaration{The authors declare no competing interest.}
\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: yuempark@berkeley.edu}

% At least three keywords are required at submission. Please provide three to five keywords, separated by the pipe symbol.
\keywords{silicate weathering $|$ weatherability $|$ arc-continent collision $|$ Neogene cooling $|$ Southeast Asian islands}

@@

Over geologic time-scales, \COtwo enters Earth's ocean--atmosphere system primarily via volcanism and metamorphic degassing, and leaves primarily through the chemical weathering of silicate rocks, with a smaller sink through organic carbon burial \cite{Kump1997a}. Chemical weathering delivers alkalinity and cations to the ocean which drives carbon sequestration through carbonate precipitation. Steady-state \pCOtwo is set at the \pCOtwo level at which \COtwo sinks are equal to sources. As \COtwo sinks are removed and \pCOtwo rises, temperature increases and the hydrological cycle is invigorated, causing other weathering sinks to increase until a new steady-state is achieved at higher \pCOtwo \cite{Kump1997a}. Conversely, as regions with high carbon sequestration potential emerge on Earth's surface, \pCOtwo decreases resulting in other weathering sinks to diminish leading to a new lower steady-state \pCOtwo.

Topography, climate, and lithology all effect chemical weathering. High-relief regions lack extensive regolith development, and thus tend to have reaction-limited weathering regimes that are more sensitive to climate change \cite{Gabet2009a, West2012a}. In warm and wet regions, mineral dissolution kinetics are faster leading to enhanced chemical weathering \cite{West2012a}. Mafic rocks have higher Ca and Mg concentrations and dissolution rates than felsic rocks, and thus the potential to more efficiently sequester carbon through silicate weathering \cite{Dessert2003a}. These factors have lead to the proposal that arc-continent collisions within the tropical rain belt have been important in enhancing global weatherability, lowering atmospheric \pCOtwo, and initiating glacial climate over the past 520~m.y. \cite{Jagoutz2016a, Swanson-Hysell2017a, Macdonald2019a} and perhaps in the Neoproterozoic as well \cite{Park2019b}.

Previous work has estimated that the decrease in \pCOtwo associated with the emergence of the SEAI since 5~Ma is $\sim$19~ppm \cite{Molnar2015a}. This 19~ppm estimate was obtained using a 0-D equation that assumes a direct linear relationship between mean global temperature and changes in weathering-rate-weighted land area, scaled by a factor that is intended to account for the influence of both runoff and temperature. \pCOtwo was then estimated from the calculated temperature using a simple energy balance equation. However, the relationship between mean global temperature (or \pCOtwo) and weathering-rate-weighted land area is not linear. As regions with high carbon sequestration potential, such as the SEAI, emerge on Earth's surface, the way in which silicate weathering rates evolve around the globe to adjust to the higher global weatherability depends on the specific spatial distribution of topography, climate, and lithology at the time of emergence. For example, if the spatial distribution of topography, climate, and lithology were such that Earth's surface overall had a relatively low carbon sequestration potential at the time of SEAI emergence, then a larger \pCOtwo change would be required to balance the sinks and sources of \COtwo than if Earth's surface overall had a relatively high carbon sequestration potential. However, silicate weathering within the emerging SEAI themselves would be adapting to changing \pCOtwo, ultimately resulting in a complex, non-linear, and spatially-varying system of feedbacks between climate and silicate weathering. Therefore, any attempt to meaningfully estimate the decrease in steady-state \pCOtwo associated with emergence of the SEAI must account for the spatial distribution of topography, climate, and lithology.

\section*{Methods}

\subsection*{GEOCLIM model}

To estimate the decrease in steady-state \pCOtwo associated with the increase of subaerially-exposed land area in the SEAI, we use the globally spatially-resolved GEOCLIM model \cite{Godderis2017c}. GEOCLIM estimates changes in steady-state \pCOtwo associated with coupled changes in erosion, chemical weathering, and climatology by linking a silicate weathering model to climate model runs at multiple \pCOtwo levels.

\begin{figure}[h]
    \centering
    \includegraphics[width=1\linewidth]{Figures/regolith_schematic.pdf}
    \caption{A schematic representation of the silicate weathering component of GEOCLIM in a single profile at steady-state. A rock particle leaves the unweathered bedrock with production rate $P_{r}$, and transits vertically through a regolith of height $h$. Regolith production and physical erosion ($E_{p}$) are equal at steady-state. As the particle transits upwards, some fraction of the primary phases ($x$) are chemically weathered ($W$) with the flux of Ca and Mg corresponding to that weathering rate multiplied by the concentration of Ca+Mg in unweathered bedrock ($\chi_{CaMg}$). Details of the formulation for the silicate weathering component of GEOCLIM can be found in \MM.}
    \label{fig:regolith_schematic}
\end{figure}

The silicate weathering component of GEOCLIM calculates \COtwo consumption resulting from silicate weathering for subaerially-exposed land. In previous versions of the model, silicate weathering was a function of temperature and runoff only, and all bedrock was assigned identical chemical compositions \cite{Godderis2017c}. More recent versions of GEOCLIM implement regolith development and soil shielding (Fig. \ref{fig:regolith_schematic}), which introduces a dependence on erosion (and therefore topographic slope) \cite{Maffre2018a}. While this introduction of regolith development into GEOCLIM is important for assessing the impact of tropical arc-continent collisions on \pCOtwo, the relatively high Ca+Mg concentration in arc rocks relative to other lithologies must also be considered. We therefore implement variable bedrock Ca+Mg concentration into GEOCLIM (\SI). The spatial distribution of lithologies is sourced from the Global Lithologic Map (GLiM) \cite{Hartmann2012a} and is represented by 6 categories: metamorphic, felsic, intermediate, mafic, carbonate, and siliciclastic sediment. Each land pixel is assigned these lithologic categories at a resolution of 0.1\degrees $\times$ 0.1\degrees. The Ca+Mg concentrations of felsic, intermediate, and mafic lithologies are assigned based on the mean of data compiled from EarthChem (\url{www.earthchem.org/portal}). Given that GLiM does not distinguish ultramafic lithologies, such rocks are grouped with mafic rocks. As a result, the Ca+Mg concentration is likely an underestimate in regions of obducted ophiolites, such that the estimated effect of these regions on changing steady-state \pCOtwo is conservative. The weathering of carbonate does not contribute to long-term \COtwo consumption and its Ca+Mg concentration is ignored. The Ca+Mg concentrations of metamorphic and siliciclastic sediment lithologies are more difficult to define, since their chemical composition is strongly dependent on protolith composition and, in the case of siliciclastic sediment, the degree of previous chemical depletion. We explore a range of feasible Ca+Mg concentrations for metamorphic rocks and siliciclastic sediment during calibration of the silicate weathering component.

\subsection*{Calibration}

The values of four parameters within the silicate weathering component that modify the dependence of silicate weathering on temperature, runoff, and regolith thickness are poorly constrained. Rather than prescribing values, we allow these parameters and the Ca+Mg concentration of metamorphic and siliciclastic pixels to vary within reasonable bounds (\SI). This results in 93,600 unique combinations of these six parameters. For each combination, we compute spatially-resolved long-term \COtwo consumption associated with Ca+Mg fluxes using present-day runoff, temperature, and slope. We sum computed \COtwo consumption over watersheds for which data-constrained estimates are available \cite{Gaillardet1999a, Moquet2018a}, then calculate the coefficient of determination ($r^{2}$) between computed and measured \COtwo consumption in each of these watersheds. After eliminating parameter combinations that result in low $r^{2}$, 5,381 parameter combinations remain. The resulting global \COtwo consumption of these model runs all overlap with independently-derived estimates of global \COtwo outgassing flux as they should with the long-term carbon cycle being at steady-state (\SI).

Having calibrated the silicate weathering component of GEOCLIM, we use it to estimate the decrease in steady-state \pCOtwo associated with SEAI emergence. For the climate model component, we use temperature and runoff from a subset of the GFDL CM2.0 experiments \cite{Delworth2006b} (\SI). The primary strength of these experiments for this analysis is that all non-\COtwo forcings are held constant at values representative of pre-industrial conditions, allowing the effect of changing \pCOtwo on climatology to be isolated.

\subsection*{Paleoshorelines}

\begin{figure*}[h]
    \centering
    \includegraphics[width=11.4cm]{Figures/shoreline_growth.pdf}
    \caption{The emergence of the Southeast Asian islands (also referred to as the Maritime Continent in climate science literature) from the mid-Miocene to present. Past shorelines are shown in \textit{A} with associated land area summarized in \textit{B}. A significant increase in area over the past 5 million years is coincident with cooling and the onset of Northern Hemisphere glaciation as reflected in the benthic oxygen isotope record \cite{Zachos2008a} shown in \textit{C}.}
    \label{fig:shoreline_growth}
\end{figure*}

To determine the position of SEAI paleoshorelines over the past 15~m.y., we use terrestrial and marine sedimentary deposits (Fig. \ref{fig:shoreline_growth}; \SI). The paleoshoreline data indicate that the Sunda-Banda Arc and New Guinea are primarily responsible for the increase in area since 15~Ma. Exhumation of the modern Sunda-Banda Arc is the result of ongoing arc-continent collision with the subducting Australian Plate \cite{Harris2006a}. Of the major islands in this arc, only Sumatra, Belitung, Bangka, Java, Bali, and Flores were subaerially exposed before 5~Ma. However, most of Sumatra and Java along with the non-volcanic islands of the Outer Banda Arc were elevated above sea level after 5~Ma \cite{Hall2013b}. In New Guinea, emergence in the Middle Miocene is associated with collision between the Melanesian Arc and Australia's distal margin \cite{Cloos2005a}, which drove exhumation of the Irian-Marum-April Ophiolite Belt. Exhumation accelerated over the past 4~m.y. in the New Guinea Central Range due to slab-breakoff and buoyant uplift, and in eastern New Guinea due to jamming of the north-dipping subduction zone \cite{Cloos2005a}. These tectonic drivers and others throughout the region lead to progressive emergence over the past 15~m.y. that accelerated following 5~Ma (Fig. \ref{fig:shoreline_growth}B). This trend mirrors broad cooling over the Neogene that resulted in the initiation of Northern Hemisphere ice sheets (Fig. \ref{fig:shoreline_growth}C). We also include changes in areas of presently-submerged continental shelves such as the Sunda Shelf that were previously exposed (\SI).

We use GEOCLIM to estimate \pCOtwo associated with the reconstructed subaerial extent of the SEAI at ca. 15, 10, and 5~Ma (``paleo-SEAI'' scenarios; Fig. \ref{fig:scenario_pCO2}). Because we use a climate model forced with modern geography, the position of the tectonic blocks remain fixed. Although there has been motion of these tectonic blocks since 15~Ma, they have remained within tropical latitudes such that this fixed scenario is a good approximation of the paleogeography (\SI). We also test an end-member scenario, in which all islands associated with arc-continent collision in the region are removed (``removed SEAI'' scenario).

\section*{Results and Discussion}

\subsection*{Neogene cooling}

\begin{figure*}[h]
    \centering
    \includegraphics[width=1\linewidth]{Figures/scenario_pCO2.pdf}
    \caption{Steady-state \pCOtwo estimates from GEOCLIM for the various scenarios discussed in the text. For each of the seven scenarios, each point represents an estimate from one of the 5,381 unique parameter combinations that resulted in reasonable total global \COtwo consumption and most closely matched estimates of present-day \COtwo consumption in 80 watersheds around the world (\SI). The box encloses the middle 50\% of the \pCOtwo estimates (i.e. the interquartile range), and the notch represents the median with its 95\% confidence interval. The whiskers extend to the 2.5 and 97.5 percentile values. Glaciation thresholds \cite{DeConto2008a} are shown on the x-axis.}
    \label{fig:scenario_pCO2}
\end{figure*}

Using the 5,381 unique parameter combinations, the ``paleo-SEAI'' scenarios resulted in 378--447~ppm \pCOtwo for 5~Ma, 435--532~ppm for 10~Ma, and 493--709~ppm for 15~Ma (Fig. \ref{fig:scenario_pCO2}). These results indicate a progressive decrease in \pCOtwo over the Neogene associated with the emergence of the SEAI, and suggest that without this emergence, pre-industrial \pCOtwo would have been $\sim$493--709~ppm. Proxy-based estimates of the magnitude and trajectory of \pCOtwo change from the Miocene to the Pliocene are variable between techniques and associated assumptions underlying their interpretation \cite{Foster2017a}. The modeled \pCOtwo values for 15~Ma resemble the higher end of proxy-based \pCOtwo estimates for the early-mid-Miocene (\SI), indicating that the increase in subaerially-exposed land area and tectonic topography of the SEAI is sufficient to explain long-term cooling of Earth's climate over the Neogene. The \pCOtwo threshold for Antarctic glaciation is estimated to be $\sim$750~ppm with that for Northern Hemisphere glaciation being significantly lower at $\sim$280~ppm \cite{DeConto2008a}. These modeled values of decreasing \pCOtwo are therefore consistent with the record of Neogene climate with Miocene ice sheets on Antarctica \cite{Sugden1995a} followed by Northern Hemisphere ice sheets developing in the Pliocene \cite{Haug2005a} as \pCOtwo subsequently decreased.

@@

\subsection*{Marine isotopes}

Himalayan uplift \cite{Raymo1992a, Galy2007a} or a decrease in volcanic outgassing \cite{Berner1983a} have also been proposed as drivers for Neogene cooling. Rising marine \SrSr has been associated with increasing weathering of the high \SrSr Himalayan lithologies \cite{Raymo1992a}. This increase in \SrSr since ca. 35~Ma is followed by a decrease in slope ca. 15~Ma that has been attributed to exhumation of relatively low \SrSr and high \OsOs Lesser Himalayan lithologies \cite{Colleps2018a}, but could also be at least partially driven by the exhumation of mafic and organic-rich forearc sediments in the SEAI during arc-continent collision. Enhanced organic matter burial in the Bay of Bengal due to high sedimentation rates could also have contributed to Neogene cooling \cite{Galy2007a}.

It has been argued that enhanced weathering in regions of active uplift predicts an increase in global alkalinity fluxes. However, global alkalinity delivery from silicate weathering needs to be nearly constant to keep the long-term carbon cycle in steady-state \cite{Kump1997a}. Enhanced weathering in a region such as the SEAI is compensated by a decrease in weathering elsewhere. Global alkalinity delivery from silicate weathering does not change, but occurs more efficiently and thereby at lower \pCOtwo. Given that carbonate weathering is disconnected from the long-term carbon-cycle mass balance, changes in carbonate accumulation through time \cite{Si2019a} could be driven by changes in carbonate weathering.

\section*{Conclusions}

Coupled geological constraints and modeling experiments demonstrate that the SEAI has been a growing hot spot for carbon sequestration from the Miocene to present. Changes in volcanic outgassing and paleogeography elsewhere on Earth, particularly in the Himalaya and Central America, would have also effected geological carbon sources and sinks. Such changes are likely to have induced complex feedbacks on many Earth surface processes. Yet, not only does the history of SEAI emergence coincide with Neogene cooling and the onset of Northern Hemisphere glaciation, but our coupled weathering-climate model also indicates that the associated steady-state \pCOtwo change is sufficient to explain the magnitude of cooling. These results highlight that the Earth's climate state is particularly sensitive to changes in tropical paleogeography.

\matmethods{
The code for the GEOCLIM model used in this study can be found at: \url{https://github.com/piermafrost/GEOCLIM-dynsoil-steady-state}. The code that generated the inputs and analyzed the output of the GEOCLIM model can be found at: \url{https://github.com/Swanson-Hysell-Group/GEOCLIM_Modern}.

\subsection*{GEOCLIM silicate weathering component}

The silicate weathering component of the GEOCLIM model has been modified from the previously published version \cite{Godderis2017b}. The new component implements the model of Gabet and Mudd (2009) \cite{Gabet2009a} for the development of a chemically weathered profile. We refer to this chemically weathered profile as regolith where the base of the regolith is unweathered bedrock. In the model of Gabet and Mudd (2009), material enters the regolith and leaves either as a solute through chemical weathering of the material during its travel from the bedrock towards the surface, or as a physically weathered particle once it reaches the top. We use the DynSoil implementation of the Gabet and Mudd (2009) model, which integrates chemical weathering within the regolith \cite{West2012a}. The transient time-varying version of this regolith model is described by three equations:

\begin{equation}
    \frac{dh}{dt} = P_{r} - E_{p}
    \label{eq:1}
\end{equation}

\begin{equation}
    \frac{\partial x}{\partial t} = -P_{r} \frac{\partial x}{\partial z} - K \tau^{\sigma}x
    \label{eq:2}
\end{equation}

\begin{equation}
    \frac{\partial \tau}{\partial t} = -P_{r} \frac{\partial \tau}{\partial z} + 1
    \label{eq:3}
\end{equation}

\noindent
Equation \ref{eq:1} is a statement of material conservation, where $h$ is the total height of the regolith (m), $t$ is the model time (yr), $P_{r}$ is the regolith production rate (m/yr), and $E_{p}$ is the physical erosion rate (m/yr). Equation \ref{eq:2} describes how the residual fraction of weatherable phases ($x$, unitless) changes as a function of time ($t$, yr) and depth ($z$, m). $K \tau^{\sigma}$ is the dissolution rate constant, which depends on the local climate (captured by $K$, yr$^{-1-\sigma}$) and the time that a given rock particle has spent in the regolith ($\tau$, yr) to some power $\sigma$ (unitless) which implements a time-dependence. Equation \ref{eq:3} describes how the time that a given rock particle has spent in the regolith changes as time in the model progresses.

The net weathering rate in the regolith column ($W$, m/yr) can then be calculated with:

\begin{equation}
    W = \int_{0}^{h} K \tau^{\sigma} x\;dz
    \label{eq:4}
\end{equation}

The regolith production rate can be expressed as the product of the optimal production rate ($P_{0}$) and a soil production function ($f(h)$):

\begin{equation}
    P_{r} = P_{0}\;f(h)
    \label{eq:5}
\end{equation}

\begin{equation}
    P_{0} = k_{rp}\;q\;e^{-\frac{E_{a}}{R}\left(\frac{1}{T}-\frac{1}{T_{0}}\right)}
    \label{eq:6}
\end{equation}

\begin{equation}
    f(h) = e^{-\frac{-h}{d_{0}}}
    \label{eq:7}
\end{equation}

\noindent
$P_{0}$ is the `optimal' regolith production rate (m/yr), which is defined to be the regolith production rate when there is no overlying regolith. In Equation \ref{eq:6}, where $k_{rp}$ is a proportionality constant (unitless), $q$ is the runoff (m/yr), $E_{a}$ is the activation energy (J/K/mol), $R$ is the ideal gas constant (J/mol), $T$ is the temperature (K), and $T_{0}$ is the reference temperature (K), we parameterize the `optimal' regolith production rate \cite{Carretier2014a}. $f(h)$ is the soil production function (unitless), which describes how regolith production decreases as the thickness of the regolith increases. It takes an exponential form as is commonly applied in the literature \cite{Gabet2009a}. In Equation \ref{eq:7}, $d_{0}$ is a reference regolith thickness (m) \cite{Heimsath1997a}.

Our implementation of the erosion rate is parameterized based on runoff and slope ($s$; m/m):

\begin{equation}
    E_{p} = k_{e}\;q^{m}\;s^{n}
    \label{eq:8}
\end{equation}

\noindent
$k_{e}$ is a proportionality constant ((m/yr)$^{1-m}$) and $m$ and $n$ are adjustable exponents that are kept as 0.5 and 1 \cite{Maffre2018a}. This formulation is directly inspired by the stream power law \cite{Davy2000a}. This formulation and these exponent values are supported by compilations, but variability in the proportionality constant is difficult to capture at a global scale \cite{Lague2013a}.

The $K$ in the dissolution rate constant in Equation \ref{eq:2} describes the dependence of the chemical weathering on climate:

\begin{equation}
    K = k_{d}\left(1-e^{-k_{w}q}\right)e^{-\frac{E_{a}}{R}\left(\frac{1}{T}-\frac{1}{T_{0}}\right)}
    \label{eq:9}
\end{equation}
